---
title: "iBiogeobears Analysis Report"
format:
  html:
    toc: true
  pdf:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
tables_dir <- normalizePath(file.path("..", "tables"), mustWork = FALSE)
logs_dir <- normalizePath(file.path("..", "logs"), mustWork = FALSE)
figures_dir <- normalizePath(file.path("..", "figures"), mustWork = FALSE)

read_table_if_exists <- function(filename) {
  path <- file.path(tables_dir, filename)
  if (!file.exists(path)) {
    return(NULL)
  }
  utils::read.csv(path, check.names = FALSE)
}

input_validation <- read_table_if_exists("input_validation.csv")
model_status <- read_table_if_exists("model_run_status.csv")
if (is.null(model_status)) {
  model_status <- read_table_if_exists("model_run_plan.csv")
}
model_comparison <- read_table_if_exists("model_comparison.csv")
model_sensitivity <- read_table_if_exists("model_sensitivity.csv")
model_parameters <- read_table_if_exists("model_parameters.csv")
node_state_summary <- read_table_if_exists("node_state_summary.csv")
node_state_sensitivity <- read_table_if_exists("node_state_sensitivity.csv")
event_summary <- read_table_if_exists("event_summary.csv")
best_fit_events <- read_table_if_exists("best_fit_events.csv")
range_change_events <- read_table_if_exists("range_change_events.csv")
bsm_run_status <- read_table_if_exists("bsm_run_status.csv")
bsm_event_summary <- read_table_if_exists("bsm_event_summary.csv")
biogeographic_process_summary <- read_table_if_exists("biogeographic_process_summary.csv")
bsm_replicate_counts <- read_table_if_exists("bsm_replicate_counts.csv")
bsm_dispersal_routes <- read_table_if_exists("bsm_dispersal_routes.csv")
bsm_events <- read_table_if_exists("bsm_events.csv")
bsm_event_times <- read_table_if_exists("bsm_event_times.csv")
root_state_probabilities <- read_table_if_exists("root_state_probabilities.csv")
workflow_manifest <- read_table_if_exists("workflow_manifest.csv")

figure_path <- function(name) {
  path <- file.path(figures_dir, paste0(name, ".png"))
  if (file.exists(path)) path else NULL
}

include_named_figure <- function(name, caption) {
  path <- figure_path(name)
  if (!is.null(path)) {
    cat("### ", caption, "\n\n", sep = "")
    print(knitr::include_graphics(path))
    cat("\n\n")
  }
  invisible(path)
}

print_table <- function(x, digits = 4) {
  if (is.null(x) || nrow(x) == 0L) {
    cat("No table was generated for this section.\n")
    return(invisible(NULL))
  }
  numeric_cols <- vapply(x, is.numeric, logical(1))
  x[numeric_cols] <- lapply(x[numeric_cols], round, digits = digits)
  knitr::kable(x)
}

answer_for <- function(summary_item) {
  if (is.null(model_sensitivity) || !"summary_item" %in% names(model_sensitivity)) {
    return(NA_character_)
  }
  value <- model_sensitivity$answer[match(summary_item, model_sensitivity$summary_item)]
  if (length(value) == 0L || is.na(value)) NA_character_ else as.character(value)
}

best_model_label <- function() {
  if (is.null(model_comparison) || nrow(model_comparison) == 0L) {
    return("Not available")
  }
  best <- model_comparison[which.min(model_comparison$AICc), , drop = FALSE]
  if ("AICc" %in% names(best)) {
    paste0(best$model[1], " (AICc ", round(best$AICc[1], 4), ")")
  } else {
    best$model[1]
  }
}

completed_model_count <- function() {
  if (is.null(model_status) || !"status" %in% names(model_status)) {
    return("Not available")
  }
  paste0(sum(model_status$status == "completed", na.rm = TRUE), " of ", nrow(model_status))
}

warning_count_total <- function() {
  if (is.null(model_status) || !"warning_count" %in% names(model_status)) {
    return("Not available")
  }
  as.character(sum(model_status$warning_count, na.rm = TRUE))
}

bsm_summary_label <- function() {
  if (is.null(bsm_run_status) || nrow(bsm_run_status) == 0L) {
    return("Not run")
  }
  completed <- sum(bsm_run_status$status %in% c("completed", "partial"), na.rm = TRUE)
  maps <- sum(bsm_run_status$completed_maps, na.rm = TRUE)
  paste0(completed, " model(s), ", maps, " completed map(s)")
}

j_caution_label <- function() {
  plus_j_near <- identical(answer_for("plus_j_models_within_delta_aicc_2"), "yes")
  best_is_j <- identical(answer_for("best_overall_is_plus_j"), "yes")
  if (plus_j_near || best_is_j) "Yes" else "No"
}
```

## Analysis Summary

This report summarizes a single-clade BioGeoBEARS workflow managed by
`iBiogeobears`. Raw BioGeoBEARS outputs are stored separately from derived
tables, figures, and reports.

```{r}
executive_summary <- data.frame(
  item = c(
    "Fitted models",
    "Best statistical model",
    "+J caution triggered",
    "BSM stochastic mapping",
    "Captured warning count",
    "Tables directory",
    "Figures directory",
    "Raw BioGeoBEARS directory"
  ),
  value = c(
    completed_model_count(),
    best_model_label(),
    j_caution_label(),
    bsm_summary_label(),
    warning_count_total(),
    tables_dir,
    figures_dir,
    normalizePath(file.path("..", "raw_biogeobears"), mustWork = FALSE)
  ),
  stringsAsFactors = FALSE
)
print_table(executive_summary)
```

## Input Validation

```{r}
print_table(input_validation)
```

## Output Manifest

```{r}
if (!is.null(workflow_manifest) && nrow(workflow_manifest) > 0L) {
  manifest_counts <- stats::aggregate(
    relative_path ~ category,
    data = workflow_manifest,
    FUN = length
  )
  names(manifest_counts) <- c("category", "file_count")
  print_table(manifest_counts)
} else {
  cat("Workflow manifest was not available for this run.\n")
}
```

## Model Run Status

### Run Overview

```{r}
status_cols <- c("model", "status", "run_action", "logLik", "num_params", "warning_count")
if (!is.null(model_status)) {
  print_table(model_status[intersect(status_cols, names(model_status))])
} else {
  print_table(NULL)
}
```

### Output Paths

```{r}
path_cols <- c("model", "raw_output_dir", "log_file", "error_message")
if (!is.null(model_status)) {
  print_table(model_status[intersect(path_cols, names(model_status))])
} else {
  print_table(NULL)
}
```

## Model Run Warnings

Warnings emitted by BioGeoBEARS and its optimizer are summarized here. Full
stdout/stderr output remains available in each model-specific log file.

```{r}
if (!is.null(model_status) && all(c("warning_count", "warning_messages") %in% names(model_status))) {
  warning_rows <- model_status[!is.na(model_status$warning_count) & model_status$warning_count > 0L, , drop = FALSE]
  if (nrow(warning_rows) > 0L) {
    print_table(warning_rows[intersect(c("model", "warning_count", "warning_messages", "log_file"), names(warning_rows))])
  } else {
    cat("No model-level warnings were captured during BioGeoBEARS execution.\n")
  }
} else {
  cat("Warning summary columns are not available for this run.\n")
}
```

## Model Comparison

Model comparison should be interpreted as evidence about statistical fit, not
as an automatic biological conclusion. Results involving `+J` founder-event
speciation models should be reported alongside non-`+J` alternatives and
sensitivity summaries.

### Fit Metrics

```{r}
model_comparison_png <- figure_path("model_comparison")
if (!is.null(model_comparison_png)) {
  knitr::include_graphics(model_comparison_png)
}
```

```{r}
fit_cols <- c("model", "model_family", "has_j", "logLik", "num_params", "AICc", "delta_aicc", "aicc_weight")
if (!is.null(model_comparison)) {
  print_table(model_comparison[intersect(fit_cols, names(model_comparison))])
} else {
  print_table(NULL)
}
```

### Interpretation Notes

```{r}
interpretation_cols <- c("model", "caution_flag", "interpretation_note")
if (!is.null(model_comparison)) {
  print_table(model_comparison[intersect(interpretation_cols, names(model_comparison))])
} else {
  print_table(NULL)
}
```

## Model Sensitivity Summary

This section summarizes `+J` sensitivity checks in plain language. It is meant
to guide interpretation, not to automatically declare a biological conclusion.

```{r}
sensitivity_cols <- c(
  "section", "display_label", "answer", "models", "model_count", "evidence",
  "interpretation_note"
)
if (!is.null(model_sensitivity)) {
  print_table(model_sensitivity[intersect(sensitivity_cols, names(model_sensitivity))])
} else {
  print_table(NULL)
}
```

## Parameter Estimates

```{r}
parameter_cols <- c("model", "parameter", "type", "is_free", "est", "min", "max", "desc")
if (!is.null(model_parameters)) {
  free_or_core <- model_parameters$is_free | model_parameters$parameter %in% c("d", "e", "j")
  parameter_rows <- model_parameters[free_or_core, intersect(parameter_cols, names(model_parameters)), drop = FALSE]
  print_table(parameter_rows)
} else {
  print_table(NULL)
}
```

## Event Summary

This section summarizes range changes inferred from highest-probability
ancestral states along branches. It is useful as a quick overview, but it is
not a stochastic mapping event-count analysis.

```{r}
event_summary_png <- figure_path("event_summary")
if (!is.null(event_summary_png)) {
  knitr::include_graphics(event_summary_png)
}
```

```{r}
event_summary_cols <- c(
  "model", "location", "event_label", "event_count", "changed_edges",
  "interpretation_note"
)
if (!is.null(event_summary)) {
  event_rows <- event_summary[order(event_summary$model, event_summary$location, -event_summary$event_count), , drop = FALSE]
  print_table(event_rows[intersect(event_summary_cols, names(event_rows))])
} else {
  print_table(NULL)
}
```

### Best-Fit Model Event Timing and Direction

The table below lists deterministic range-change events under the best-fitting
statistical model. Times are approximate branch-midpoint values in the input
tree's branch-length units before present. They are not stochastic mapping event
times.

```{r}
best_fit_event_cols <- c(
  "event_index", "model", "event_time_midpoint", "event_time_min",
  "event_time_max", "direction_label", "direction", "event_label",
  "parent_state", "child_state", "node_label", "interpretation_note"
)
if (!is.null(best_fit_events)) {
  event_rows <- best_fit_events
  if ("event_time_midpoint" %in% names(event_rows)) {
    event_rows <- event_rows[order(-event_rows$event_time_midpoint), , drop = FALSE]
  }
  print_table(event_rows[intersect(best_fit_event_cols, names(event_rows))])
} else {
  print_table(NULL)
}
```

### Biogeographical Stochastic Mapping (BSM)

This section reports formal BioGeoBEARS stochastic mapping output when
`analysis$run_stochastic_mapping` was enabled. These tables should be preferred
over deterministic best-state changes when discussing event counts, event
timing, and source-to-target dispersal directions.

#### Biogeographic process synthesis

The synthesis below translates BSM event codes into named biogeographic
processes, split into cladogenetic speciation modes (in-situ/sympatric
speciation, subset sympatry, vicariance, and founder-event jump speciation) and
anagenetic range changes (range expansion, local extinction, and range
switching). Proportions are reported within each process group and across all
processes, so the relative importance of each process can be read directly.

```{r}
process_synthesis_png <- figure_path("biogeographic_process_synthesis")
if (!is.null(process_synthesis_png)) {
  knitr::include_graphics(process_synthesis_png)
}
```

```{r}
process_cols <- c(
  "model", "process_group", "process_label", "mean_count", "sd_count",
  "proportion_within_group", "proportion_overall"
)
if (!is.null(biogeographic_process_summary) && nrow(biogeographic_process_summary) > 0) {
  print_table(biogeographic_process_summary[intersect(process_cols, names(biogeographic_process_summary))])
} else {
  cat("Process synthesis is available only when BSM stochastic mapping was run.\n")
}
```

```{r}
if (!is.null(bsm_run_status)) {
  bsm_status_cols <- c(
    "model", "status", "requested_maps", "completed_maps",
    "maxtries_per_branch", "seed", "log_file", "error_message"
  )
  print_table(bsm_run_status[intersect(bsm_status_cols, names(bsm_run_status))])
} else {
  cat("BSM stochastic mapping was not run for this workflow.\n")
}
```

```{r}
bsm_event_summary_png <- figure_path("bsm_event_summary")
if (!is.null(bsm_event_summary_png)) {
  knitr::include_graphics(bsm_event_summary_png)
}
```

```{r}
bsm_summary_cols <- c(
  "model", "event_label", "mean_count", "sd_count", "sum_count",
  "replicate_count", "interpretation_note"
)
if (!is.null(bsm_event_summary)) {
  bsm_rows <- bsm_event_summary[order(bsm_event_summary$model, -bsm_event_summary$mean_count), , drop = FALSE]
  print_table(bsm_rows[intersect(bsm_summary_cols, names(bsm_rows))])
} else {
  print_table(NULL)
}
```

```{r}
bsm_routes_png <- figure_path("bsm_dispersal_routes")
if (!is.null(bsm_routes_png)) {
  knitr::include_graphics(bsm_routes_png)
}
```

```{r}
bsm_route_cols <- c(
  "model", "route_type", "direction_label", "mean_count", "sd_count",
  "interpretation_note"
)
if (!is.null(bsm_dispersal_routes)) {
  route_rows <- bsm_dispersal_routes[
    !is.na(bsm_dispersal_routes$mean_count) & bsm_dispersal_routes$mean_count > 0,
    ,
    drop = FALSE
  ]
  route_rows <- route_rows[order(route_rows$model, route_rows$route_type, -route_rows$mean_count), , drop = FALSE]
  print_table(utils::head(route_rows[intersect(bsm_route_cols, names(route_rows))], 30L))
} else {
  print_table(NULL)
}
```

```{r}
bsm_times_png <- figure_path("bsm_event_times")
if (!is.null(bsm_times_png)) {
  knitr::include_graphics(bsm_times_png)
}
```

```{r}
bsm_time_cols <- c(
  "model", "replicate", "event_class", "event_label",
  "event_time_before_present", "direction_label", "parent_state",
  "child_state", "node_label"
)
if (!is.null(bsm_event_times)) {
  time_rows <- bsm_event_times[order(bsm_event_times$model, bsm_event_times$replicate, -bsm_event_times$event_time_before_present), , drop = FALSE]
  print_table(utils::head(time_rows[intersect(bsm_time_cols, names(time_rows))], 50L))
} else {
  print_table(NULL)
}
```

## Root State Probabilities

```{r}
if (!is.null(root_state_probabilities)) {
  root_rows <- root_state_probabilities[order(root_state_probabilities$model, -root_state_probabilities$probability), , drop = FALSE]
  root_rows <- utils::head(root_rows, 12L)
  print_table(root_rows)
} else {
  print_table(NULL)
}
```

```{r}
root_state_png <- figure_path("root_state_probabilities")
if (!is.null(root_state_png)) {
  knitr::include_graphics(root_state_png)
}
```

## Node State Summary

This table reports the highest-probability range state per model, node, and
BioGeoBEARS probability location. It is a compact node-level interface for
downstream plotting and future Shiny views.

```{r}
included_node_figures <- c(
  !is.null(include_named_figure("node_state_summary_best_model", "Best Statistical Model")),
  !is.null(include_named_figure("node_state_summary_best_non_j", "Best Non-+J Model")),
  !is.null(include_named_figure("node_state_summary_best_plus_j", "Best +J Model"))
)
if (!any(included_node_figures)) {
  fallback_node_state_png <- figure_path("node_state_summary")
  if (!is.null(fallback_node_state_png)) {
    knitr::include_graphics(fallback_node_state_png)
  }
}
```

```{r}
node_summary_cols <- c(
  "model", "location", "node_index", "node_type", "node_label",
  "best_state", "best_probability", "state_count"
)
if (!is.null(node_state_summary)) {
  node_rows <- node_state_summary[order(node_state_summary$model, node_state_summary$location, node_state_summary$node_index), , drop = FALSE]
  node_rows <- utils::head(node_rows, 20L)
  print_table(node_rows[intersect(node_summary_cols, names(node_rows))])
} else {
  print_table(NULL)
}
```

## Node State Sensitivity

This table compares the best non-`+J` model with the best `+J` model at each
node. Rows with different best states or large probability differences are the
most useful places to inspect biological sensitivity.

```{r}
node_sensitivity_png <- figure_path("node_state_sensitivity")
if (!is.null(node_sensitivity_png)) {
  knitr::include_graphics(node_sensitivity_png)
}
```

```{r}
node_sensitivity_cols <- c(
  "location", "node_index", "node_type", "node_label",
  "non_j_model", "non_j_state", "non_j_probability",
  "plus_j_model", "plus_j_state", "plus_j_probability",
  "state_differs", "probability_difference_abs"
)
if (!is.null(node_state_sensitivity) && nrow(node_state_sensitivity) > 0L) {
  sensitivity_rows <- node_state_sensitivity[
    order(-node_state_sensitivity$state_differs, -node_state_sensitivity$probability_difference_abs),
    ,
    drop = FALSE
  ]
  sensitivity_rows <- utils::head(sensitivity_rows, 20L)
  print_table(sensitivity_rows[intersect(node_sensitivity_cols, names(sensitivity_rows))])
} else {
  cat("Node-state sensitivity comparison is not available because both +J and non-+J completed models are required.\n")
}
```

## Methodological Cautions

`iBiogeobears` keeps the following cautions visible by default:

- report all fitted models, not only the minimum-AIC model;
- compare `+J` and non-`+J` model families explicitly;
- separate statistical model fit from biological interpretation;
- report uncertainty and stochastic mapping summaries where available;
- discuss whether founder-event speciation dominates inferred events.

```{r}
if (!is.null(model_comparison) && "caution_flag" %in% names(model_comparison)) {
  caution_rows <- model_comparison[model_comparison$caution_flag != "none", , drop = FALSE]
  if (nrow(caution_rows) > 0L) {
    print_table(caution_rows[intersect(c("model", "delta_aicc", "caution_flag", "interpretation_note"), names(caution_rows))])
  } else {
    cat("No model-specific caution flags were triggered beyond the default interpretation notes.\n")
  }
} else {
  cat("Model comparison output is not available.\n")
}
```

## Software and Citation

BioGeoBEARS is authored by Nicholas J. Matzke and must be cited when used.
In R, run:

```r
citation("BioGeoBEARS")
```

`iBiogeobears` does not bundle BioGeoBEARS. It checks for an external
BioGeoBEARS installation and records package versions and session information in
the workflow output directory.

```{r}
citation_path <- file.path(logs_dir, "biogeobears_citation.txt")
if (file.exists(citation_path)) {
  citation_lines <- readLines(citation_path, warn = FALSE)
  if (length(citation_lines) > 0L && any(nzchar(citation_lines))) {
    cat(paste(citation_lines, collapse = "\n"))
  } else {
    cat("BioGeoBEARS citation metadata was not available in this run.")
  }
} else {
  cat("BioGeoBEARS citation metadata was not available in this run.")
}
```
